---
title: 'Template: Mapping Canada'
author: "Marco Lugo"
date: "February 4, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen=999) # avoid scientific notation (useful for legend)

library(RColorBrewer)
library(maptools)
library(ggmap)
library(scales)
```

## Purpose

This is a quick example on how to create a map in R using shapefiles provided by Statistics Canada. For the purpose of this example we have also collected population density data for map visualization.

- The shapefiles used here can be obtained from: http://www12.statcan.gc.ca/census-recensement/2011/geo/bound-limit/bound-limit-2011-eng.cfm

- Population data (2011 Census) can be downloaded here: http://www12.statcan.gc.ca/census-recensement/2011/dp-pd/hlt-fst/pd-pl/FullFile.cfm?T=1201&LANG=Eng&OFT=CSV&OFN=98-310-XWE2011002-1201.CSV


## Load Data

### Load Population Data

```{r load_data1}
df_pop <- read.csv('98-310-XWE2011002-1201.CSV')
colnames(df_pop)
head(df_pop)

# only keep the Forward Sortation Area (FSA; first three characters of the Postal Code) and the pop count
df_pop <- subset(df_pop, select = c('Geographic.name', 'Population..2011'))
# Rename variables
colnames(df_pop) <- c('fsa', 'Population')

# drop the total for Canada as a whole, we are only interested in FSAs
df_pop <- subset(df_pop, fsa != 'Canada')
head(df_pop)
```

### Load Canada FSA Shapefiles

```{r load_data2}
can_sf <- readShapePoly('FSA/gfsa000b11a_e.shp')
head(can_sf@data$CFSAUID)
```

## Prepare Base Map

```{r map_1}
# In Ubuntu, you may need to run the following command before installing packages rgeos, rgdal and gpclib, which are needed for the fortify command below:
# sudo apt-get install libgdal1-dev libproj-dev

can_df <- fortify(can_sf, region = 'CFSAUID')
rm(can_sf) # free up memory

# get map from Google Maps API
gmap <- qmap("Canada", zoom=3)

# build polygon boundaries from shapefile data
fsa_boundaries <- geom_polygon(aes(x=long, y=lat, group=group), data=can_df, color='black', fill=NA)

# let's have a look. no population for now
gmap + fsa_boundaries
```

The previous map, while it covered all of Canada, may not be ideal for observing FSA-level differences. Let's zoom in.

```{r map_2}
gmap <- qmap("Montreal, Canada", zoom=12)
gmap + fsa_boundaries
```

## Merge Data

```{r merge_data1}
# add population counts to map data
can_df <- merge(can_df, df_pop, by.x='id', by.y='fsa', all.x=TRUE)
```

## Population Map

```{r map_3}
fsa_boundaries_pop <- geom_polygon(aes(x=long, y=lat, group=group, fill=Population, alpha=0.2), data=can_df, color='black')
gmap + fsa_boundaries_pop + scale_fill_gradient(low='#ffeda0', high='#f03b20') + scale_alpha(guide='none')
```

This map is using a population scale that is difficult to understand, let's redesign it.

```{r new_scale1}
quantile(df_pop$Population, probs=seq(0.0, 1, 0.1), na.rm=TRUE)
```

We see that 90% of Canadian FSAs have ~40k or below. Let's create new categories that reflect this.

```{r new_scales2}
new_breaks <- seq(from=0, to=40000, by=10000)
new_labels <- comma(new_breaks)
new_labels[length(new_labels)] <- paste0(new_labels[length(new_labels)], '+') # append plus sign for the highest label
new_labels

gmap + fsa_boundaries_pop +
  scale_fill_gradient(low='#ffeda0', high='#f03b20', limits=c(0, 40000), breaks=new_breaks, labels=new_labels) + 
  scale_alpha(guide='none')
```

